First Season MWA EoR Power Spectrum Results at Redshift 7

### 2016-06-02 ms.tex
Addressing Caths comments, updating a couple references

Using the first generation of instruments, several upper limits have been placed on the 21 
cm EoR signal \citep{Ali:2015, Dillon:2014, Parsons:2014, Jacobs:2015, Paciga:2013}, and 
\citealt{Pober:2015} was able to place constraints on physical reionization models. 
However, much work is to be done to understand the data produced by these arrays. 
Foreground subtraction and isolation has emerged as a priority in the field. Central to most 
analysis strategies is the concept of an ``EoR window" -- the region of Fourier space where 
spectrally smooth foregrounds have been isolated from the isotropic (spherically symmetric 
in Fourier space) cosmological signal \citep{Morales:2006, Bowman:2009}. More recent 
studies have shown the existence of a foreground ``wedge", the result of instrumental mode 
mixing, throwing power from spectrally smooth foregrounds to higher Fourier modes 
\citep{Thyagarajan:2015b, Thyagarajan:2015, Trott:2012, Liu:2014a, Liu:2014b, 
Hazelton:2013, Pober:2013, Vedantham:2012, Morales:2012, Datta:2010}. Despite this 
setback, the EoR window is preserved above the wedge, and the field is pushing forward with 
several analysis pipelines under active development \citep[e.g. B.~J.~Hazelton et al. 2016, in 
prep; D.~A.~Mitchell et al. 2016, in prep; D.~C.~Jacobs et al. 2016, in review; ][]{Trott:2016, 
Dillon:2013, Trott:2014}.

@@

The remainder of this paper is organized as follows. In Section~\ref{sec:MWA} we briefly 
describe the MWA instrument and the observations used in this work, in Section~
\ref{sec:analysis} we describe our analysis pipeline, in Section~\ref{sec:techniques} we 
describe several novel techniques in our analysis, in Section~\ref{sec:deep} we discuss our 
efforts to apply the analysis pipeline to a deep data integration \textbf{and compare with a reference pipeline for robust estimation}, and we discuss future work 
in Section~\ref{sec:discussion}. Throughout this paper we use a $\Lambda$CDM 
cosmology with $\Omega_m=0.73$, $\Omega_\Lambda=0.27$, and $h = 0.7$, consistent 
with WMAP seven year results \citep{Komatsu:2011}. All distances and wavenumbers are in 
comoving coordinates.

@@

The technical design of the MWA is reviewed in \citealt{Tingay:2013}, and we highlight a 
few key characteristics that will become important in our analysis of the data. Each 
antenna of the MWA comprises 16 dual-polarization dipoles placed on a regular grid, lying 
on a ground screen. The radio frequency signals from these dipoles feed into an analog 
beamformer which uses physical delays to ``point" the antenna. The MWA contains 128 of 
these antennas, with a tightly packed core of radius 50 m, and extending out to a radius \textbf{of}
1.5 km for higher resolution calibration and imaging.

@@

\section{Analysis Pipeline}\label{sec:analysis}
In order to ensure consistency in analysis, the international MWA EoR collaboration has 
defined two independent reference analysis pipelines. The details of this strategy are 
outlined in D.~C.~Jacobs et al. 2016 (in review). This study is based on the FHD-to-
\eppsilon pipe, which is in turn based on the Fast Holographic Deconvolution (FHD
\footnote{\url{https://github.com/EoRImaging/FHD}}, \citealt{Sullivan:2012}) and Error 
Propagated Power Spectrum with Interleaved Observed Noise (\eppsilon
\footnote{\url{https://github.com/EoRImaging/eppsilon}}, B.~J.~Hazelton et al. 2016, in 
prep) packages. The general flow of the pipeline is shown in Figure~\ref{fig:pipe}, and we 
give an overview below. \textbf{As a point of comparison, and to demonstrate robustness of the results, we compute the output power spectra with an independent calibration and power spectrum estimation pipeline, and demonstrate consistency.}

@@

\subsection{Preprocessing}\label{subsec:preprocessing}
The preprocessing step converts the data from the non-standard memory dump format 
from the correlator's GPUs to a standard UVFITS format \citep{Greisen:2012}, performs 
flagging, averages in time and freqeuncy, and writes to disk. For this step we use the 
\cotter package, which in turn calls the \texttt{AOFLAGGER}\footnote{\url{http://
aoflagger.sourceforge.net/}} to perform RFI flagging \citep{Offringa:2010}.

Besides RFI, we manually flag data to remove three other effects. As mentioned in 
Section~\ref{sec:MWA}, the edges of the coarse bands contain high levels of aliasing from 
the filter applied in the digital receivers. We flag two 40 kHz channels on either side of the 
coarse band edges (total of four channels per coarse band edge). In principle the aliasing 
could be calibrated out and these channels could be recovered, but this is left for future 
work. Second, early data from the MWA also contained some instances where antennas 
were not pointed properly at the beginning of observations due to a beamformer error. This 
problem has since been resolved, but exists in our data. We therefore flag the first two 
seconds of each observation to avoid any potentially mis-pointed data. Finally, we flag the 
central fine channel of each coarse band because it corresponds to the DC mode of that 
sub-band and can easily float, washing out useful information. \textbf{Please reword this sentence. It is not clear what "float" means here. Can you be more specific?}

@@

\section{Novel Analysis Methods}\label{sec:techniques}
\textcolor{red}{Needs a better section title}
\textbf{Analysis Methods for Testing Data Quality}

@@

Because we use the input foreground model to calibrate our visibilities, the gain estimates 
for our two runs differed. In particular we saw the overall flux scale of the calibrated 
visibilities was higher when using the full hierarchical catalog. In order to put both residual 
2D power spectra onto the same scale for comparison, we first divide each pixel by the 
corresponding dirty power spectrum pixel, then subtract to arrive at a ratio difference. 
\begin{equation}\label{eq:diff_ratio}
\mathtt{ratio\;difference = \frac{residual_1}{dirty_1} - \frac{residual_2}{dirty_2}}
\end{equation}
Here we use subscript 1 to represent the MWACS catalog spectra, and 2 for the 
hierarchical catalog spectra. The ratio difference is shown in Figure~\ref{fig:mwacs_ratio}. 
The entire wedge being positive (blue) demonstrates that using the new catalog resulted in 
successfully subtracting higher fractional power. Not only does this indicate we subtract 
more power, but it also confirms that our calibrated data is more closely matched to the 
model, meaning that our calibration solutions in general are more accurate. \textbf{Consider adding a reference and a sentence about Jack's paper here. Her performed quite a few of these type of tests for understanding the impact of having an incomplete point source model.}

@@

@@

We demonstrate the impact of our diffuse model again by running the golden data set 
through our entire analysis pipeline with and without using the diffuse model. Because the 
foreground models used for calibration in the two runs are identical (diffuse is omitted for 
calibration), the calibration solutions were identical and therefore the dirty power spectra 
were identical as well. We compare the model and residual power spectra by direct 
subtraction in Figure~\ref{fig:diffuse_diff}. The top panel shows the model power with the 
diffuse subtracted from the model power without. The entire plot is negative (red) because 
we only added power with the diffuse model. The bottom panel is the difference in residual 
power. The wedge is completely positive (blue), indicating that the diffuse model 
successfully subtracted from the dirty visibilities. The differences in the window are positive 
and negative, indicating that any differences are below the three hour noise level.   \textbf{Why is the region in the EoR window red here, particularly when the lower plot shows noise-like differences?}

@@

As demonstrated in D.~C.~Jacobs et al. 2016 (in review), comparison between independent 
analysis pipelines is essential to evaluate the validity of algorithms under active development. 
Such comparison provides insight into the strengths and weaknesses of the pipelines, as well 
as confirmation of results. The analysis discussed so far has been based on the FHD to 
\eppsilon pipeline, and here we compare with the Real Time System \citep[RTS,][]
{Mitchell:2008,Ord:2010} to Cosmological \ion{H}{1} Power Spectrum \citep[CHIPS,][]
{Trott:2016} pipeline. Below we present a brief overview of the reference pipeline and 
highlight lessons learned from the comparison.

>>>>

Using the first generation of instruments, several upper limits have been placed on the 21 
cm EoR signal \citep{Ali:2015, Dillon:2014, Parsons:2014, Jacobs:2015, Paciga:2013}, and 
\citealt{Pober:2015} was able to place constraints on physical reionization models. 
However, much work is to be done to understand the data produced by these arrays. 
Foreground subtraction and isolation has emerged as a priority in the field. Central to most 
analysis strategies is the concept of an ``EoR window" -- the region of Fourier space where 
spectrally smooth foregrounds have been isolated from the isotropic (spherically symmetric 
in Fourier space) cosmological signal \citep{Morales:2006, Bowman:2009}. More recent 
studies have shown the existence of a foreground ``wedge", the result of instrumental mode 
mixing, throwing power from spectrally smooth foregrounds to higher Fourier modes 
\citep{Thyagarajan:2015b, Thyagarajan:2015, Trott:2012, Liu:2014a, Liu:2014b, 
Hazelton:2013, Pober:2013, Vedantham:2012, Morales:2012, Datta:2010}. Despite this 
setback, the EoR window is preserved above the wedge, and the field is pushing forward with 
several analysis pipelines under active development \citep[e.g. B.~J.~Hazelton et al. 2016, in 
prep; D.~A.~Mitchell et al. 2016, in prep; ][]{Jacobs:2016, Trott:2016, 
Dillon:2013, Trott:2014}.

@@

The remainder of this paper is organized as follows. In Section~\ref{sec:MWA} we briefly 
describe the MWA instrument and the observations used in this work, in Section~
\ref{sec:analysis} we describe our analysis pipeline, in Section~\ref{sec:techniques} we 
describe several novel techniques in our analysis, in Section~\ref{sec:deep} we discuss our 
efforts to apply the analysis pipeline to a deep data integration and compare with a reference pipeline for robust estimation, and we discuss future work 
in Section~\ref{sec:discussion}. Throughout this paper we use a $\Lambda$CDM 
cosmology with $\Omega_m=0.73$, $\Omega_\Lambda=0.27$, and $h = 0.7$, consistent 
with WMAP seven year results \citep{Komatsu:2011}. All distances and wavenumbers are in 
comoving coordinates.

@@

The technical design of the MWA is reviewed in \citealt{Tingay:2013}, and we highlight a 
few key characteristics that will become important in our analysis of the data. Each 
antenna of the MWA comprises 16 dual-polarization dipoles placed on a regular grid, lying 
on a ground screen. The radio frequency signals from these dipoles feed into an analog 
beamformer which uses physical delays to ``point" the antenna. The MWA contains 128 of 
these antennas, with a tightly packed core of radius 50 m, and extending out to a radius of
1.5 km for higher resolution calibration and imaging.

@@

\section{Analysis Pipeline}\label{sec:analysis}
In order to ensure consistency in analysis, the international MWA EoR collaboration has 
defined two independent reference analysis pipelines. The details of this strategy are 
outlined in \citealt{Jacobs:2016}. This study is based on the FHD-to-
\eppsilon pipe, which is in turn based on the Fast Holographic Deconvolution 
(FHD\footnote{\url{https://github.com/EoRImaging/FHD}}, \citealt{Sullivan:2012}) and Error 
Propagated Power Spectrum with Interleaved Observed Noise (\eppsilon
\footnote{\url{https://github.com/EoRImaging/eppsilon}}, B.~J.~Hazelton et al. 2016, in 
prep) packages. The general flow of the pipeline is shown in Figure~\ref{fig:pipe}, and we 
give an overview below. As a point of comparison, and to demonstrate robustness of the results, we compute the output power spectra with an independent calibration and power spectrum estimation pipeline, and demonstrate consistency.

@@

\subsection{Preprocessing}\label{subsec:preprocessing}
The preprocessing step converts the data from the non-standard memory dump format 
from the correlator's GPUs to a standard UVFITS format \citep{Greisen:2012}, performs 
flagging, averages in time and freqeuncy, and writes to disk. For this step we use the 
\cotter package \citep{Offringa:2015}, which in turn calls the \texttt{AOFLAGGER}\footnote{\url{http://
aoflagger.sourceforge.net/}} to perform RFI flagging \citep{Offringa:2010}.

Besides RFI, we manually flag data to remove three other effects. As mentioned in 
Section~\ref{sec:MWA}, the edges of the coarse bands contain high levels of aliasing from 
the filter applied in the digital receivers. We flag two 40 kHz channels on either side of the 
coarse band edges (total of four channels per coarse band edge). In principle the aliasing 
could be calibrated out and these channels could be recovered, but this is left for future 
work. Second, early data from the MWA also contained some instances where antennas 
were not pointed properly at the beginning of observations due to a beamformer error. This 
problem has since been resolved, but exists in our data. We therefore flag the first two 
seconds of each observation to avoid any potentially mis-pointed data. 
Finally, we flag the central fine channel of each coarse band. This channel corresponds
to the sub-band DC mode, which has been observed to contain anomalous power likely
due to very low level rounding errors in the polyphase filter bank of the digital receivers.

@@

\section{Analysis Methods for Testing Data Quality}\label{sec:techniques}

@@

Because we use the input foreground model to calibrate our visibilities, the gain estimates 
for our two runs differed. In particular we saw the overall flux scale of the calibrated 
visibilities was higher when using the full hierarchical catalog. In order to put both residual 
2D power spectra onto the same scale for comparison, we first divide each pixel by the 
corresponding dirty power spectrum pixel, then subtract to arrive at a ratio difference. 
\begin{equation}\label{eq:diff_ratio}
\mathtt{ratio\;difference = \frac{residual_1}{dirty_1} - \frac{residual_2}{dirty_2}}
\end{equation}
Here we use subscript 1 to represent the MWACS catalog spectra, and 2 for the 
hierarchical catalog spectra. The ratio difference is shown in Figure~\ref{fig:mwacs_ratio}. 
The entire wedge being positive (blue) demonstrates that using the new catalog resulted in 
successfully subtracting higher fractional power. Not only does this indicate we subtract 
more power, but it also confirms that our calibrated data is more closely matched to the 
model, meaning that our calibration solutions in general are more accurate.

@@

J.~L.~B.~Line, et al (2016, in review) performed similar analysis, but with an emphasis on the
positional accuracy of catalog entries. By using simulations of visibilities they were able to 
isolate the effect of using perfect versus offset source positions. They confirmed the
importance of a complete and accurate point source catalog in order to retain a clean EoR
window.

@@

We demonstrate the impact of our diffuse model again by running the golden data set 
through our entire analysis pipeline with and without using the diffuse model. Because the 
foreground models used for calibration in the two runs are identical (diffuse is omitted for 
calibration), the calibration solutions were identical and therefore the dirty power spectra 
were identical as well. We compare the model and residual power spectra by direct 
subtraction in Figure~\ref{fig:diffuse_diff}. The top panel shows the model power with the 
diffuse subtracted from the model power without. The entire plot is negative (red) because 
we only added power with the diffuse model. 
Even the EoR window is red (albeit at a much lower level than the wedge) because of power
leakage from non-uniform spectral sampling and from the dynamic range limit of the 
Blackman-Harris window function.
The bottom panel is the difference in residual 
power. The wedge is completely positive (blue), indicating that the diffuse model 
successfully subtracted from the dirty visibilities. The differences in the window are positive 
and negative, indicating that 
the differences in the foreground model in this region are below the three hour noise level.

@@

As demonstrated in \citealt{Jacobs:2016}, comparison between independent 
analysis pipelines is essential to evaluate the validity of algorithms under active development. 
Such comparison provides insight into the strengths and weaknesses of the pipelines, as well 
as confirmation of results. The analysis discussed so far has been based on the FHD to 
\eppsilon pipeline, and here we compare with the Real Time System \citep[RTS,][]
{Mitchell:2008,Ord:2010} to Cosmological \ion{H}{1} Power Spectrum \citep[CHIPS,][]
{Trott:2016} pipeline. Below we present a brief overview of the reference pipeline and 
highlight lessons learned from the comparison.
